---
title: "Results"
bibliography: references.bib
---

```{r setup-results, include=FALSE, message=FALSE, warning=FALSE}
library(plotly)

source("scripts/sex_comparison_analysis copy.R")
source("scripts/Fresh_KEGG_Analysis copy.R")
```

# Results

## Key Findings

- DCM samples show stronger gene expression differences than sex alone.
- Sex differences are present but more noticeable in DCM samples.
- PCA and UMAP show partial clustering between groups.
- KEGG analysis highlights pathways related to inflammation and metabolism.

---

## Visualizations

## PCA Plot - Sex Comparison

```{r pca-plot, echo=FALSE, message=FALSE, warning=FALSE}
plot_ly(
  x = pca$x[,1],
  y = pca$x[,2],
  type = "scatter",
  mode = "markers",
  color = meta_sex$Group,
  colors = group_colors,
  text = paste(
    "Sample:", meta_sex$Sample,
    "<br>Group:", meta_sex$Group,
    "<br>Sex:", meta_sex$Sex,
    "<br>Condition:", meta_sex$Condition
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "PCA Plot - Sex Comparison",
    xaxis = list(title = paste0("PC1 (", pc1_var, "%)")),
    yaxis = list(title = paste0("PC2 (", pc2_var, "%)"))
  )
```

This PCA plot compares control and DCM samples by sex. Each point represents one sample. The plot shows some separation between disease groups, but there is still overlap, meaning sex-specific differences are present but not the only factor affecting gene expression.

Beyond simply showing separation between groups, the PCA results provide insight into the structure of variation within the dataset. The fact that PC1 captures the largest proportion of variance and primarily reflects disease status suggests that DCM induces widespread transcriptional remodeling across the genome. This aligns with prior RNA-seq studies showing that large numbers of genes can be differentially expressed in DCM, reflecting broad biological disruption rather than isolated pathway changes.

The lack of strong separation by sex along the principal components further suggests that sex-related transcriptional differences are subtle relative to disease-driven effects. However, the overlap between groups also indicates that DCM is not a uniform condition, and variability within DCM samples may reflect differences in disease stage, underlying cause, or patient-specific factors.

---

## UMAP Plot - Sex Comparison

```{r umap-plot, echo=FALSE, message=FALSE, warning=FALSE}
plot_ly(
  x = ump$layout[,1],
  y = ump$layout[,2],
  type = "scatter",
  mode = "markers",
  color = meta_sex$Group,
  colors = group_colors,
  text = paste(
    "Sample:", meta_sex$Sample,
    "<br>Group:", meta_sex$Group,
    "<br>Sex:", meta_sex$Sex,
    "<br>Condition:", meta_sex$Condition
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "UMAP Plot - Sex Comparison",
    xaxis = list(title = "UMAP1"),
    yaxis = list(title = "UMAP2")
  )
```

This UMAP plot provides another view of sample clustering. It shows nonlinear patterns in the data that PCA may not capture as clearly. The groups are not perfectly separated, which suggests that multiple biological factors are influencing the expression patterns.

While PCA captures linear variance, the UMAP visualization highlights the complexity of gene expression relationships that are not easily explained by a single axis of variation. The partial clustering observed in UMAP suggests that DCM samples share common transcriptional signatures but still retain heterogeneity.

This heterogeneity is consistent with current understanding of DCM as a multifactorial disease involving diverse molecular mechanisms, including inflammation, fibrosis, and metabolic remodeling. Importantly, UMAP reveals that gene expression differences may exist along gradients rather than discrete categories, reinforcing the idea that disease progression occurs along a continuum rather than a binary state.

---

## Volcano Plot - Female vs Male DCM

```{r volcano-dcm-plot, echo=FALSE, message=FALSE, warning=FALSE}
plot_ly(
  data = volcano_dcm_df,
  x = ~logFC,
  y = ~plotNegLogP,
  type = "scatter",
  mode = "markers",
  color = ~Significance,
  colors = c(
    "Female Higher" = "deeppink",
    "Male Higher" = "blue",
    "Not Significant" = "grey"
  ),
  text = paste(
    "Gene:", rownames(volcano_dcm_df),
    "<br>logFC:", round(volcano_dcm_df$logFC, 3),
    "<br>P-value:", signif(volcano_dcm_df$P.Value, 3),
    "<br>Status:", volcano_dcm_df$Significance
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Volcano Plot - Female vs Male DCM",
    xaxis = list(title = "log Fold Change (Female vs Male)"),
    yaxis = list(title = "-log10 P-value")
  )
```

This volcano plot compares female and male DCM samples. Points farther left or right show stronger sex-based expression differences. Colored points represent genes that are higher in females or higher in males, suggesting that DCM may affect gene expression differently by sex.

The differential expression observed between male and female DCM samples suggests that sex may influence how the heart responds to pathological stress. Prior studies have shown that inflammatory signaling pathways and immune cell activity can differ between males and females, potentially contributing to differences in disease severity and progression.

Additionally, many of the differentially expressed genes likely reflect processes such as extracellular matrix remodeling and cytokine signaling, which are hallmarks of cardiac dysfunction. These findings support the idea that sex-specific regulatory mechanisms may influence the balance between adaptive and maladaptive remodeling in DCM.

---

## Volcano Plot - Female vs Male Control

```{r volcano-control-plot, echo=FALSE, message=FALSE, warning=FALSE}
plot_ly(
  data = volcano_control_df,
  x = ~logFC,
  y = ~plotNegLogP,
  type = "scatter",
  mode = "markers",
  color = ~Significance,
  colors = c(
    "Female Higher" = "deeppink",
    "Male Higher" = "blue",
    "Not Significant" = "grey"
  ),
  text = paste(
    "Gene:", rownames(volcano_control_df),
    "<br>logFC:", round(volcano_control_df$logFC, 3),
    "<br>P-value:", signif(volcano_control_df$P.Value, 3),
    "<br>Status:", volcano_control_df$Significance
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Volcano Plot - Female vs Male Control",
    xaxis = list(title = "log Fold Change (Female vs Male)"),
    yaxis = list(title = "-log10 P-value")
  )
```

This volcano plot compares female and male control samples. Compared to the DCM volcano plot, fewer strong differences are expected here. This supports the idea that sex-based gene expression differences may become more noticeable in disease.

The relatively limited number of differentially expressed genes between males and females in control samples suggests that baseline transcriptional differences are modest in healthy tissue. This reinforces the idea that sex differences become more biologically relevant under stress or disease conditions.

This pattern has been observed in other cardiovascular studies, where sex-based differences in gene expression are often amplified during disease progression rather than present at baseline.

---

## PC1 Distribution

```{r pc1-plot, echo=FALSE, message=FALSE, warning=FALSE}
plot_ly(
  data = pc1_df,
  x = ~Group,
  y = ~PC1,
  color = ~Group,
  colors = group_colors,
  type = "box",
  boxpoints = "all",
  jitter = 0.3,
  pointpos = 0
) %>%
  layout(
    title = "PC1 Distribution by Group",
    xaxis = list(title = ""),
    yaxis = list(title = "PC1 Score")
  )
```

This boxplot shows how PC1 values are distributed across each group. PC1 represents the largest source of variation in the dataset. Differences between groups suggest that disease status and sex may contribute to major gene expression variation.

The distribution of PC1 values across groups further supports the conclusion that disease status is the dominant factor influencing gene expression. The shift in PC1 values in DCM samples reflects coordinated changes across multiple genes and pathways, rather than isolated gene-level effects.

This type of coordinated shift is characteristic of large-scale biological processes such as metabolic reprogramming and inflammatory activation, both of which are known to play central roles in DCM progression.

---

## KEGG Pathway Analysis

KEGG enrichment analysis was performed on significantly differentially expressed genes to understand biological pathways involved in sex differences in DCM.

## KEGG Summary Plot - Female DCM vs Male DCM

```{r kegg-plot, echo=FALSE, message=FALSE, warning=FALSE}
invisible(capture.output(
  suppressWarnings(
    source("scripts/Fresh_KEGG_Analysis copy.R")
  )
))

if ("P.Value" %in% colnames(results) && nrow(results) > 0) {
  
  kegg_plot_df <- results[order(results$P.Value), ]
  kegg_plot_df <- head(kegg_plot_df, 15)
  
  kegg_plot_df$GeneRatio <- as.numeric(kegg_plot_df$GeneRatio)
  kegg_plot_df$SignificantGenesInPathway <- as.numeric(kegg_plot_df$SignificantGenesInPathway)
  kegg_plot_df$adj.P.Val <- as.numeric(kegg_plot_df$adj.P.Val)
  kegg_plot_df$PathwayName <- as.character(kegg_plot_df$PathwayName)
  
  plot_ly(
    data = kegg_plot_df,
    x = ~GeneRatio,
    y = ~PathwayName,
    type = "scatter",
    mode = "markers",
    size = ~SignificantGenesInPathway,
    color = ~adj.P.Val,
    colors = "RdBu",
    text = paste(
      "Pathway:", kegg_plot_df$PathwayName,
      "<br>Gene Ratio:", round(kegg_plot_df$GeneRatio, 3),
      "<br>Significant Genes:", kegg_plot_df$SignificantGenesInPathway,
      "<br>P-value:", signif(kegg_plot_df$P.Value, 3),
      "<br>Adjusted P-value:", signif(kegg_plot_df$adj.P.Val, 3)
    ),
    hoverinfo = "text"
  ) %>%
    layout(
      title = "KEGG Pathway Enrichment: Female DCM vs Male DCM",
      xaxis = list(title = "Gene Ratio"),
      yaxis = list(title = "KEGG Pathway")
    )
  
} else {
  cat("No KEGG pathways were available to plot.")
}
```

This interactive KEGG plot shows the top enriched pathways from the Female DCM vs Male DCM comparison. Larger points represent pathways with more significant genes, and the color represents the adjusted p-value. This helps connect gene expression differences to larger biological pathways.

The KEGG pathway enrichment results provide critical biological context for the observed gene expression changes. The enrichment of inflammatory and metabolic pathways is consistent with previous transcriptomic studies of DCM, which have identified shifts in energy metabolism, mitochondrial function, and immune signaling as key features of the disease.

For example, metabolic reprogramming in DCM often involves a shift away from oxidative phosphorylation toward alternative energy pathways, reflecting impaired mitochondrial function. At the same time, increased inflammatory signaling and cytokine activity contribute to fibrosis and structural remodeling of the heart.

The presence of these pathways in the current analysis reinforces the biological validity of the results and demonstrates that the observed gene expression differences are not random, but instead reflect well-established mechanisms of disease progression.

---

## Interpretation

Overall, the results demonstrate that dilated cardiomyopathy is associated with widespread transcriptional changes that affect multiple biological systems simultaneously. Disease status consistently emerged as the primary driver of variation across all analyses, indicating that DCM induces large-scale remodeling of gene expression.

While sex-based differences were not dominant at the global level, they became more apparent within the DCM group, suggesting that disease conditions may amplify underlying biological differences between males and females. This supports the idea that sex is an important modifying factor rather than a primary driver of gene expression differences.

Importantly, pathway analysis revealed that these differences are concentrated in biologically meaningful processes, including inflammation, metabolism, and cardiac remodeling. These findings align with existing literature and reinforce the role of these pathways in DCM progression.

Together, these results support the hypothesis that DCM alters gene expression in a sex-dependent manner, although the magnitude of these effects is smaller than the overall impact of disease.

---

## Discussion


At a mechanistic level, the observed gene expression changes likely reflect the combined effects of cardiomyocyte stress, immune activation, and structural remodeling. Previous studies have shown that DCM progression involves activation of profibrotic signaling pathways, including TGF-beta signaling, as well as shifts in metabolic gene expression that impair energy production in cardiac cells [@rosenkranz2015; @burchfield2013; @heinig2017]. These processes contribute to decreased contractile function and progressive heart failure.

The enrichment of metabolic pathways in this study is particularly important, as metabolic dysfunction is increasingly recognized as a central feature of cardiomyopathy. Alterations in pathways such as oxidative phosphorylation and fatty acid metabolism indicate that the failing heart undergoes significant energy remodeling, which may contribute to disease progression and symptom severity [@seemann2017; @zhao2020; @smith2021].

Furthermore, the identification of sex-specific differences within these pathways suggests that males and females may experience distinct molecular responses to the same disease condition. This has important implications for personalized medicine, as treatments targeting metabolic or inflammatory pathways may have different effects depending on sex [@kararigas2018; @raimondi2020].

---

## Biological Significance


The findings from this study highlight the importance of integrating transcriptomic data with biological interpretation. Rather than representing isolated gene-level changes, the observed differences reflect coordinated shifts in biological systems that govern cardiac function.

Understanding these patterns is critical for identifying potential therapeutic targets, as interventions aimed at single genes may be less effective than those targeting entire pathways. By identifying key pathways associated with DCM and sex-specific differences, this study contributes to a broader understanding of the molecular basis of heart failure (@alimadadi2020; @yu2012).

---

## Limitations

This study has several limitations. The analysis uses bulk RNA-seq data, meaning that gene expression changes cannot be assigned to specific cardiac cell types. Patient-level variability, including disease stage, treatment history, genetic background, and clinical severity, may also influence gene expression patterns. In addition, although pathway analysis provides useful biological context, some pathways may not reach strong statistical significance after correction for multiple testing. Therefore, these results should be interpreted as exploratory and would benefit from validation in additional datasets.

---

## Future Work

Future work should include larger and more diverse patient cohorts to improve the detection of subtle sex-specific gene expression differences. Integrating additional omics approaches, such as proteomics, epigenomics, or single-cell RNA-seq, could help identify which cell types and molecular mechanisms contribute most strongly to DCM progression. Experimental validation of key genes and pathways would also strengthen the biological relevance of the findings and support future therapeutic research.
